---
title: "Case Study 3"
subtitle: "AKSTA Statistical Computing"
author: Elias Anton Rozanics, Stefan Sick, Filip Cernak (Group 90)
date: "2026-05-06"
output: pdf_document
---

*The  .Rmd* **and** *.html (or .pdf) should be uploaded in TUWEL by the deadline. Refrain from using explanatory comments in the R code chunks but write them as text instead. Points will be deducted if the submitted file is not in a decent form.*

```{r, warning=FALSE}
packages <- c("tidyverse")

install_if_missing <- packages[!(packages %in% installed.packages()[,"Package"])]
if(length(install_if_missing)) install.packages(install_if_missing)


library(tidyverse)
library(readxl)
library(ggplot2)
library(scales)

```

# Data

Load the data set you exported in the final Task of Case Study 2. 
Eliminate all
observations with missing values in the income status variable.

As a reminder, the data set includes world data from 2020, focusing on:

- **Education Expenditure (% of GDP)**
- **Youth Unemployment Rate (15-24 years)**
- **Net Migration Rate** (difference between the number of people entering and leaving a country per 1,000 persons)

for most world entities in 2020. The data was downloaded from https://www.cia.gov/the-world-factbook/about/archives/.
Additional information on continent, subcontinent/region and income status was appended 
to the dataset in Case Study 2.
```{r}
df_test <- read_delim("processed_data.csv", delim = ";")
 
df_all <- read_delim("processed_data.csv", delim = ";") %>% 
  mutate(ed_expenditure = as.numeric(ed_expenditure),
         youth_unempl_rate = as.numeric(youth_unempl_rate),
         continent = case_when(country_name == "Kosovo" ~ "Europe",
                               TRUE ~ continent),
         region = case_when(country_name == "Kosovo" ~ "Southern Europe",
                     TRUE ~ region),
        income_group = na_if(income_group, "."),
        income_group = str_trim(income_group), 
        income_group = fct_relevel(income_group, "High income", "Upper middle income", "Lower middle income", "Low income")) 
   
```

# Tasks:

## a. Education expenditure in different income levels

Using **ggplot2**, create a density plot of the education expenditure grouped by income 
status. The densities for the different groups are superimposed in the 
same plot rather than in different plots. Ensure that you order the levels of the 
income status such that in the plots the legend is ordered from High (H) to Low (L).

  * The color of the density lines is black.
  
  * The area under the density curve should be colored differently among the income status levels.
  
  * For the colors, choose a transparency level of 0.5 for better visibility.
  
  * Position the legend at the top center of the plot and give it no title (hint: use `element_blank()`).
  
  * Rename the x axis as "Education expenditure (% of GDP)"

  Comment briefly on the plot.


```{r}
unique(df_all$income_group)
```

```{r}

# the order of the income factors has already been set in the data preparation
df_all %>%

  filter(!is.na(income_group)) %>% 
  ggplot(aes(x = ed_expenditure, fill = income_group, group = income_group)) +
  geom_density(color = "black", alpha = 0.5) +
  labs(x = "Education expenditure (% of GDP)", 
       y = "Density") +
  theme_minimal() +
  scale_x_continuous(breaks = breaks_pretty(n = 5)) + 
  theme(
    legend.position = "top",
    legend.justification = "center",
    legend.title = element_blank()
  )
```
 
From the plot we can see that low income countries spend the least on education as their density is concentrated the most at the lower end (around 1-3% of GDP). As income increases, the distributions shift to the right, with high income countries having a peak at around 5% of GDP. Upper middle income countries show the tallest peak, which suggests a more consistent spending level across this group. Overall, higher income is generally associated with higher education expenditure, even if the distributions overlap substantially across all groups.

## b. Income status in different continents

Investigate how the income status is distributed in the different continents. 

* Using **ggplot2**, create a stacked barplot of absolute frequencies showing how the entities are split into continents and income status. Comment the plot.

* Create another stacked barplot of relative frequencies (height of the bars should be one).
Comment the plot.

* Create a mosaic plot of continents and income status using base R functions. 

* Briefly comment on the differences between the three plots generated to investigate the income 
distribution among the different continents.


```{r}
df_all %>% 
  filter(!is.na(income_group)) %>% 
  ggplot(aes(x = continent, fill = income_group, group = income_group)) +
  geom_bar(position = "stack") + 
  scale_fill_brewer(palette = "RdYlGn", direction = -1) +
  theme_minimal() 

df_all %>% 
  filter(!is.na(income_group)) %>% 
  ggplot(aes(x = continent, fill = income_group, group = income_group)) +
  geom_bar(position = "fill") + 
  scale_fill_brewer(palette = "RdYlGn", direction = -1)  + 
  labs(y = "relative frequency") + 
  theme_minimal() 


mosaic_data <- table(
  df_all %>% filter(!is.na(income_group)) %>% pull(continent),
  df_all %>% filter(!is.na(income_group)) %>% pull(income_group)
)

mosaicplot(mosaic_data, 
           main = "Mosaic Plot of Income vs Continent",
           xlab = "Continent", 
           ylab = "Income Group",
           las = 1,
           color = terrain.colors(4)) 
```
From the stacked barplot of absolute frequencies, we can see that Africa has the most countries overall and is dominated by low and lower middle income countries. Europe has a similar count of countries but almost all of them are high income. Asia and Americas show a more mixed distribution across income groups, with Asia having more lower middle income countries and Americas more high income countries. Oceania has the fewest countries with most of them being high income.

The second plot is a stacked barplot of relative frequencies, where now the height of the bars is 1, so we can do proportional comparisons across continents. Europe has the highest proportion of high income countries while Africa has the highest proportion of low income countries. The Americas are mainly high income and upper middle income countries. Asia has the most balanced distribution across all income groups.

The mosaic plot shows the same income proportions as the relative barplot, but the column widths now also reflect the total number of countries per continent. Africa is the widest column confirming it has the most countries, while Oceania is the narrowest.

Overall, the absolute barplot is best for comparing total counts. The relative barplot allows us to make proportional comparisons but we lose the count information. The mosaic plot combines both which makes it the most informative of the three.

## c. Income status in different subcontinents

For Oceania, investigate further how the income status distribution is in the different subcontinents.
Use one of the plots in b. for this purpose. Comment on the results.
```{r}

df_all %>% 
  filter(continent == "Oceania") %>% 
  distinct(region)

df_oceania <- df_all %>% 
  filter(continent == "Oceania", !is.na(income_group))

mosaic_data <- table(df_oceania$region, df_oceania$income_group)
mosaicplot(mosaic_data, 
           main = "Mosaic Plot of Income vs Subcontinent",
           xlab = "Subcontinent", 
           ylab = "Income Group",
           las = 2,
           color = terrain.colors(4)) 

```

To investigate the income status distribution in the different subcontinents of Oceania, we used a mosaic plot. New Zealand/Australia is entirely composed of high income countries but also has the fewest countries as it is the narrowest. Micronesia has the most countries, with most of them being high income. Melanesia contains the most lower middle income countries, while Polynesia contains a mix of high and upper middle income countries. This shows that the regions containing the island states contain more diversity than the New Zealand/Australia group.

## d. Net migration in different continents

* Using **ggplot2**, create parallel boxplots  showing  the distribution of the
net migration rate in the different continents. 

* Prettify the plot (change y-, x-axis labels, etc).

* Identify which country in Asia constitutes the largest negative outlier and 
which country in Asia constitutes the largest positive outlier. 

* Comment on the plot.
```{r}
df_all %>% 
  filter(!is.na(mig_rate)) %>% 
  ggplot(aes(x=continent, y = mig_rate, fill = continent)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.5) +
  geom_boxplot(outlier.color = "red", alpha= 0.7) +
  labs(x = "Continent",
       y = "Net Migration Rate",
       title = "Net Migration Rate per Continent") +
  theme_minimal()
  

# Largest negative outlier in Asia
df_all %>% 
  filter(continent == "Asia") %>% 
  slice_min(mig_rate, n = 1) %>% 
  select(country_name, mig_rate)

# Largest positive outlier in Asia
df_all %>% 
  filter(continent == "Asia") %>% 
  slice_max(mig_rate, n = 1) %>% 
  select(country_name, mig_rate)
```

From the plot we can see that most continents have median net migration rates close to 0. Europe is the only country with a positive median, confirming it as a net immigration destination. Oceania shows the most variability while also having the largest negative median. Asia, Americas and Africa all have a very slightly negative median, with Asia having the most extreme outliers. Lebanon is the largest negative outlier with a rate of -88.7 and Syria is the largest positive outlier with a rate of 27.1. The boxes for most continents are very narrow which suggests low variability in net migration rates.

## e. Net migration in different subcontinents

The graph in d. clearly does not convey the whole picture. It would be interesting also 
to look at the subcontinents, as it is likely that a lot of migration flows
happen within the continent.

* Investigate the net migration in different subcontinents
using again parallel boxplots.  Group the boxplots by continent (hint: use `facet_grid` with `scales = "free_x"`).

* Remember to prettify the plot (rotate axis labels if needed).

* Describe what you see.
```{r}
df_all %>% 
  filter(!is.na(mig_rate)) %>% 
  ggplot(aes(x = mig_rate, y = region, fill = continent)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray50", linewidth = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.5) +
  geom_boxplot(outlier.color = "red", alpha = 0.7) +
  labs(x = "Net Migration Rate (per 1,000)",
       y = "Subcontinent",
       title = "Net Migration Rate per Subcontinent") +
  facet_grid(continent ~ ., scales = "free_y") +
  theme_minimal()
```
Breaking down the plot by subcontinents reveals patterns that were not seen at the continent scale. In Africa, both subregions have median close to 0 with Sub-Saharan Africa showing much more outliers. In the Americas, Northern America shows much more variability with an overall positive median compared to Latin America and the Caribbean which has a slight negative median and much more outliers. Within Asia, Western Asia has the most variability with the most extreme outliers (Lebanon and Syria). In Europe, all subregions have a positive median, which confirms that Europe is a net immigration destination across all its subregions. In Oceania, Polynesia and Micronesia both show large variability with a negative median, which reflects the emigration from the Pacific island nations, while Australia and New Zealand have a positive median.


## f. Median net migration rate per subcontinent.

The plot in task e. shows the distribution of the net migration rate for each 
subcontinent. Here you will work on visualizing only one summary statistic, namely 
the median. 

For each subcontinent, calculate the median net migration rate. Then create a plot which
contains the sub-regions on the y-axis and the median net migration rate on 
the x-axis. 

  * As geoms use points. 
  
  *  Color the points by continent -- use a colorblind friendly palette (see e.g., 
  [here](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/)).
  
  * Rename the axes.
  
  * Using `fct_reorder` from the **forcats** package, arrange the levels of subcontinent
  such that in the plot the lowest (bottom) subcontinent contains the lowest median
  net migration
  rate and the upper most region contains the highest median
  net migration
  rate.
  
  * Comment on the plot. E.g., what are the regions with the most influx? 
  What are the regions with the most outflux?
```{r}
df_all %>%
  filter(!is.na(mig_rate), !is.na(region)) %>%
  group_by(continent, region) %>%
  summarize(median_mig = median(mig_rate, na.rm = TRUE), .groups = "drop") %>%
  mutate(region = fct_reorder(region, median_mig)) %>%
  ggplot(aes(x = median_mig, y = region, color = continent)) +
  geom_point(size = 4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray60") +
  scale_color_viridis_d(option = "viridis") + 
  labs(
    title = "Median Net Migration Rate by Sub-Region",
    x = "Median Net Migration Rate",
    y = "Sub-region",
    color = "Continent"
  ) +
  theme_minimal()
```

From the plot, we can see that the region Australia and New Zealand has the highest median net migration rate, followed by Western and Northern Europe, which makes these the regions with the most influx. On the other hand, Micronesia and Polynesia have the most negative medians, which makes them the regions with the most outflux. This reflects the emigration from Pacific island nations. Most other subregions have medians close to zero, indicating roughly balanced migration flows.

## g. Median youth unemployment rate per subcontinent

For each subcontinent, calculate the median youth unemployment rate. Then create a plot which
contains the sub-regions on the y-axis and the median unemployment rate on 
the x-axis. 

  * Use a black and white theme (`?theme_bw()`)
  
  * As geoms use bars. (hint: pay attention to the statistical transformation taking place
  in `geom_bar()` -- look into  argument `stat="identity"`) 
  
  * Color the bars by continent -- use a colorblind friendly palette.
  
  * Make the bars transparent (use `alpha = 0.7`).
  
  * Rename the axes.
  
  * Using `fct_reorder` from the **forcats** package, arrange the levels of subcontinent
  such that in the plot the lowest (bottom) subcontinent contains the lowest median
  youth unemployment rate and the upper most region contains the highest median
  youth unemployment rate.
  
  * Comment on the plot. E.g., what are the regions with the highest vs lowest youth
  unemployment rate?
  
```{r}
df_all %>%
  filter(!is.na(youth_unempl_rate), !is.na(region)) %>%
  group_by(continent, region) %>%
  summarize(median_empl = median(youth_unempl_rate, na.rm = TRUE), .groups = "drop") %>%
  mutate(region = fct_reorder(region, median_empl)) %>%
  ggplot(aes(x = median_empl, y = region, fill = continent)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  scale_fill_viridis_d(option = "viridis") + 
  labs(
    title = "Median Youth Unemployment Rate by Sub-Region",
    x = "Median Youth Unemployment Rate",
    y = "Sub-Region",
    color = "Continent"
  ) +
  theme_bw()
```

Northern Africa has the highest median youth unemployment rate (35%), followed by Polynesia and Southern Europe. Eastern Asia has the lowest median youth unemployment rate (8-9%), followed by Central Asia and South-eastern Asia. Europe shows mixed results as Southern Europe is among the highest while Northern Europe and Western Europe are closer to the middle. Most of the subregions have a median between 10-15%.

## h. Median youth unemployment rate per subcontinent -- with error bars

The value displayed in the barplot in g. is the result of an aggregation, 
so it might be useful to also plot error bars, to have a general idea on
how precise the median unemployment is. This can be achieved by plotting the
error bars which reflect the standard deviation or the interquartile range of 
the variable in each of the subcontinents. 

Repeat the plot from Task g. but include also error bars which reflect the 25% and 75%
quantiles. You can use `geom_errorbar` in **ggplot2**.


```{r}
df_all %>%
  filter(!is.na(youth_unempl_rate), !is.na(region)) %>%
  group_by(continent, region) %>%

  summarize(
    median_unempl = median(youth_unempl_rate, na.rm = TRUE),
    q1 = quantile(youth_unempl_rate, 0.25, na.rm = TRUE),
    q3 = quantile(youth_unempl_rate, 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(region = fct_reorder(region, median_unempl)) %>%
  
  ggplot(aes(x = median_unempl, y = region, fill = continent)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  geom_errorbar(aes(xmin = q1, xmax = q3), 
                width = 0.4,       
                color = "gray30") + 
  
  scale_fill_viridis_d(option = "viridis") + 
  labs(
    title = "Youth Unemployment: Median and IQR by Region",
    subtitle = "Error bars represent the Interquartile Range",
    x = "Youth Unemployment Rate (%)",
    y = "Region",
    fill = "Continent"
  ) +
  theme_bw()
```
In this plot we also included error bars which reflect the 25% and 75% quantiles, which add information about the spread in each subregion. Polynesia and Western Asia have the widest IQRs, which suggests high variability in youth unemployment rates. Northern Africa, which has the highest median, has a relatively narrow IQR, which suggests consistent high youth unemployment rates. Regions with low median like Eastern Asia and Central Asia also have narrow IQRs, which indicates consistent low youth unemployment rates.


## i. Relationship between education expenditure and net migration rate

Using **ggplot2**, create a plot showing the relationship between education expenditure and
net migration rate. 

  * Color the geoms based on the income status. 

  * Add a regression line for each development status (using `geom_smooth()`).
  
Comment on the plot. Do you see any relationship between the two variables? Do you see
any difference among the income levels?
```{r}
# One outlier in mig_rate has been removed, to improve readability of the plot

df_all %>%
  filter(!is.na(ed_expenditure), !is.na(mig_rate), !is.na(income_group), mig_rate > -25 & mig_rate < 25) %>%

  ggplot(aes(x = ed_expenditure, y = mig_rate, color = income_group)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_viridis_d(option = "viridis") +
  labs(
    title = "Relationship Between Education Expenditure and Net Migration",
    subtitle = "Fitted with linear regression lines per income group",
    x = "Education Expenditure ",
    y = "Net Migration Rate",
    color = "Income Status"
  ) +
  theme_minimal()
```

From the plot, we can see that there is no strong overall relationship between education expenditure and net migration rate. High income countries have a slight negative slope but they maintain positive migration rates across all education expenditure levels. Lower middle income countries show the steepest negative slope, which means that high education expenditure is associated with less net migration. Upper middle income and low income lines are relatively flat, indicating little relationship between the two variables in these groups. Overall, the quite wide scatter of points across all groups suggests that education expenditure is not the best predictor for net migration rate.

## j. Relationship between youth unemployment and net migration rate

Create a plot as in Task i. but for youth unemployment and net migration rate. Comment briefly.

```{r}
df_all %>%
  filter(!is.na(youth_unempl_rate), !is.na(mig_rate), !is.na(income_group),
         mig_rate > -25 & mig_rate < 25) %>%
  ggplot(aes(x = youth_unempl_rate, y = mig_rate, color = income_group)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_viridis_d(option = "viridis") +
  labs(
    title = "Relationship Between Youth Unemployment and Net Migration",
    subtitle = "Fitted with linear regression lines per income group",
    x = "Youth Unemployment Rate (%)",
    y = "Net Migration Rate (per 1,000)",
    color = "Income Status"
  ) +
  theme_minimal()
```
When looking at the plot, we can see that the regression lines differ across income groups. High income and Lower middle income countries both show a similar negative slope, meaning that higher youth unemployment is associated with lower (more negative) net migration rates. On the other hand, Upper middle income and Low income countries show a slight positive slope. But overall, no strong consistent pattern exists across all groups, and also the quite wide scatter of points suggests youth unemployment rate alone is not the best predictor for the net migration rate.


## k. Merging population data

Go online and find a data set which contains the 2020 population for the countries of the world
together with ISO codes. 

* Download this data and merge it to the dataset you are working on in
this case study using a left join.
(A possible source: [World Bank](https://data.worldbank.org/indicator/SP.POP.TOTL?end=2020&start=2020)))

* Inspect the data and check whether the join worked well.

```{r}
#https://data.worldbank.org/indicator/SP.POP.TOTL?end=2020&start=2020

pop_raw <- read_csv("population_data.csv", skip = 4)

pop_2020 <- pop_raw %>%
  select(iso_code = `Country Code`, population = `2020`) %>%
  filter(!is.na(population))

df_merged <- df_all %>%
  left_join(pop_2020, by = "iso_code")

# We check the join
cat("Rows before join:", nrow(df_all), "\n")
cat("Rows after join:", nrow(df_merged), "\n")

df_merged %>%
  filter(is.na(population)) %>%
  select(country_name, iso_code)

str(df_merged)
```

We can see that the number of rows remains the same (227 rows) before and after the join, so no duplicate matches appeared. We can also see that 10 countries have missing population data after the join, which are all small territories for which the World Bank does not have the population data.  


## l. Scatterplot of education expenditure and net migration rate in Europe 

Make a scatterplot of education expenditure and net migration rate for the countries of Europe. 

  * Scale the size of the points according to each country's population. 
 
  * For better visibility, use a transparency of `alpha=0.7`. 
  
  * Remove the legend.
  
  * Comment on the plot.
  
```{r}
df_merged %>%
  filter(continent == "Europe",
         !is.na(ed_expenditure), !is.na(mig_rate), !is.na(population)) %>%
  ggplot(aes(x = ed_expenditure, y = mig_rate,
             size = population, label = country_name)) +
  geom_point(alpha = 0.7, color = "steelblue") +
  geom_text(size = 2.5, vjust = -0.9, check_overlap = TRUE) +
  scale_size_continuous(range = c(2, 14)) +
  labs(
    title = "Education Expenditure vs. Net Migration Rate in Europe (2020)",
    x = "Education Expenditure",
    y = "Net Migration Rate"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

We can observe that there is no clear linear relationship between education expenditure and net migration rate in Europe. The biggest countries in terms of population like Germany, France and Ukraine all cluster around an education expenditure of 4-5% with near 0 to slightly positive migration rates. The highest migration rates are in Luxembourg and Monaco, which are both very small and wealthy countries, but which do not have a high education expenditure, with Monaco even having the lowest. We can also see that Nordic countries like Sweden, Norway and Denmark stand out in the plot, as they have the highest education expenditure (around 8%) and also have a high net migration rate. On the other hand, countries like Moldova, Latvia and Lithuania show a quite large negative migration rate while having an education expenditure of around 4-5%. Overall, high education expenditure does not directly drive net migration rates, as other factors like wages or geographic location may play a role.


## m. Interactive plot

On the merged data set from Task k., using function `ggplotly` from package **plotly**  
re-create the scatterplot in Task l., but this time for all countries. Color the points
according to their continent.

When hovering over the points the name of the country, 
the values for education expenditure, net migration rate, and population 
should be shown. (Hint: use the aesthetic `text = Country`. 
In `ggplotly` use the argument `tooltip = c("text", "x", "y", "size")`).

**Hint:** If you are creating a .pdf file, keep in mind that PDF output is static and does not support HTML widgets. 
To avoid build errors, you can wrap interactive plots in a conditional check 
or set the chunk option `eval=FALSE`. Alternatively, you can use the **webshot** 
package to automatically convert interactive elements into static images during the knit process.

## n. Parallel coordinate plot

In **parallel coordinate plots** each observation or data point is depicted as a line 
traversing a series of parallel axes, corresponding to a specific variable or dimension.
It is often used for identifying clusters in the data.

One can create such a plot using the **GGally** R package. You should create 
such a plot where you look at the three main variables in the data set: 
education expenditure, youth unemployment rate and net migration rate. 
Color the lines based on the income status. Briefly comment.

## o. World map visualisation

Create a world map of the education expenditure per country.
You can use the vignette https://cran.r-project.org/web/packages/rworldmap/vignettes/rworldmap.pdf
to find how to do this in R. Alternatively, you can use other packages (such as **ggplot2**, **sf** and **rnaturalearthdata**) to create a map.

