---
title: "Case Study 1"
subtitle: "AKSTA Statistical Computing"
output: pdf_document
header-includes:
  - \usepackage{booktabs}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*The  .Rmd and .html (alternatively the .pdf) should be uploaded in TUWEL by the deadline. Refrain from using explanatory comments in the R code chunks but write them as text instead. Please consider the AI Guidelines as specified in the organizational slides (you can use the table). Points will be deducted if the submitted file is not in a decent form; answers should be answering the concrete questions and not discuss topics in general - please be mindful of our tutors' time and the size of the course. *


\begin{table}[ht]
\centering
\caption{You can use and extend this table to specify your AI usage (if used at all).}
\begin{tabular}{l|rrr}
\toprule
 \textbf{AI usage} & \textbf{Task 1} & \textbf{Task 2} & \textbf{Task 5} \textbf{Task 6} \ldots \\
\midrule
\textit{Correcting faulty code}   &  no  & no &  no & little &  \\
\textit{Improving existing code}   &  no   &  no &  no & no &   \\
\textit{Generating new code}   &  no   & no &  no & no &  \\
\textit{Layout and formatting}  &  no   & no &  no & no &   \\
\textit{Text correction}   &  little  & no &  no & no &  \\
\textit{Help with Latex expressions}  &  no  & little &    \\
\textit{Explaination of R base function}  &  no & no & little & no & \\
\bottomrule
\end{tabular}
\end{table}


# 1. Ratio of Fibonacci numbers (1 point)

## a. 
Write two different R functions which return the sequence $r_{i}=F_{i+1}/F_i$
for $i=1,\ldots,n$ where $F_i$ is the $i$th Fibonacci number, once using `for`
and once using `while`.

```{r}
#for loop

fibonacci_for <- function(n) {
  F <- numeric(n + 1)
  F[1] <- 1
  F[2] <- 1
  
  for(i in 3:(n+1)){
    F[i] <- F[i-1] + F[i-2]
  }
  
  r <- numeric(n)
  for (i in 1:n) {
    r[i] <- F[i + 1] / F[i]
  }
  
  return(r)
}

#while loop

fibonacci_while <- function(n) {
  F <- numeric(n + 1)
  F[1] <- 1
  F[2] <- 1
  
  i <- 3
  while (i <= n + 1) {
    F[i] <- F[i - 1] + F[i - 2]
    i <- i + 1
  }
  
  r <- numeric(n)
  i <- 1
  while (i <= n) {
    r[i] <- F[i + 1] / F[i]
    i <- i + 1
  }
  
  return(r)
}

fibonacci_for(10)
fibonacci_while(10)
```

We can see that both of the functions give us the same results.

## b. 
Benchmark the two functions for $n=200$ and $n=2000$ (you can use package **microbenchmark** or package **bench** for this purpose). Which function is faster?

```{r}
library(microbenchmark)

#benchmark for 200

microbenchmark(
  for_function_200 = fibonacci_for(200),
  while_function_200 = fibonacci_while(200)
)

#benchmark for 2000

microbenchmark(
  for_function_2000 = fibonacci_for(2000),
  while_function_2000 = fibonacci_while(2000)
)

```

In this case, the microbenchmark() function runs by default each expression 100 times. We can observe that in both cases (n=200 and n=2000), the function using for is faster. The median time for the function using for is around 1.5 times faster than the function using while for both n=200 and n=2000. We mainly look at the median here instead of the mean as some of the max values are way too large compared to the median, and so this could have an effect on the mean.

## c.
Plot the sequence for  $n=100$. 
For which $n$ does it start to stabilize to a value?
What number does the sequence converge to?

```{r}
ratio_100 <- fibonacci_for(100)

plot(ratio_100, type = "l",
     xlab = "i",
     ylab = "r_i",
     main = "Fibonacci ratios")
```

When looking at the plot, we can see that the sequence starts to stabilize at around n=7 or n=8. For larger n, the plot looks completely flat, which shows that the sequence converges. The sequence converges towards a value around 1.62 ($\approx$ 1.61803 to be exact). This number is also called the Golden Ratio, which is a well known limit of the ratio of consecutive Fibonacci numbers.


# 2. Gamma function (1 point)

## a. 
Write a function to compute the following for $n$ a positive integer
using the `gamma` function in base R.
$$
\rho_n =\frac{\Gamma((n-1)/2)}{\Gamma(1/2)\Gamma((n - 2)/2)}
$$


```{r}
rho_n <- function(n){
  gamma((n - 1)/2) / (gamma(1/2) * gamma((n - 2)/2))
}
```

## b. 

Try $n = 2000$. What do you observe? Why do you think the observed behavior happens?

```{r}
rho_n(2000)
```

When evaluating our function for n=2000, we get NaN (not a number). This means that we have an undefined operation that is going on. Since the arguments $\frac{n-1}{2}$ and $\frac{n-2}{2}$ are very large for n=2000, and the Gamma function grows even faster than exponentially, we have in this case Inf/Inf, which gives us NaN. This happens because the values become so large before the division in both the numerator and denominator that R cannot represent them and so stores the values as Inf.

## c.
Write an implementation which can also deal with large values of $n>1000$. 

To deal with large values of n $>$ 1000, we will work with the logarithm of gamma using the function lgamma() and then exponentiate the final result.

```{r}
rho_large_n <- function(n) {
  exp(lgamma((n - 1)/2) - lgamma(1/2) - lgamma((n - 2)/2))
}

rho_large_n(2000)
```

We can now see that the stabilized function with the logarithm and exponentiation gives us a finite result even for very large n. And so instead of working with the very large Gamma values and then dividing (which leads to Inf/Inf), we work in the log space. Because log(Gamma) grows much slower than Gamma, the values we obtain are manageable for R to work with. We then subtract these log-gamma values (instead of dividing in the original formula) and then exponentiate, which gives us a finite result.

## d. 
Plot $\rho_n/\sqrt{n}$ for different values of $n$. Can you guess the limit of $\rho_n/\sqrt{n}$ for $n \rightarrow\infty$?

```{r}
n_values <- 2:500
rho_values <- sapply(n_values, rho_large_n)
plot(n_values, rho_values / sqrt(n_values), type = "l", col = "blue",
     main = expression(rho[n]/sqrt(n)),
     xlab = "n", ylab = expression(rho[n]/sqrt(n)))
abline(h = 1/sqrt(2*pi), col = "red", lty = 2)
```

To estimate the limit of 
$$
\frac{\rho_n}{\sqrt{n}} \quad \text{for } n \to \infty,
$$

let's have a closer look at the gamma functions from $\rho_n$. We know that $\Gamma\left(\frac{1}{2}\right) = \sqrt{\pi}$. For the remainder of the Gamma terms, we will use Stirling's approximation formula: 
$$
\Gamma(x) \approx \sqrt{2\pi}\, x^{x - \frac{1}{2}} e^{-x}
$$

When we apply this to our 2 remaining Gamma terms in $\rho_n$, we get:
$$
\Gamma\left(\frac{n-1}{2}\right)
\approx
\sqrt{2\pi}\left(\frac{n-1}{2}\right)^{\frac{n-1}{2} - \frac{1}{2}}
e^{-\frac{n-1}{2}}
=
\sqrt{\frac{4\pi}{n-1}}
\left(\frac{n-1}{2e}\right)^{\frac{n-1}{2}}
$$
and 
$$
\Gamma\left(\frac{n-2}{2}\right)
\approx
\sqrt{2\pi}\left(\frac{n-2}{2}\right)^{\frac{n-2}{2} - \frac{1}{2}}
e^{-\frac{n-2}{2}}
=
\sqrt{\frac{4\pi}{n-2}}
\left(\frac{n-2}{2e}\right)^{\frac{n-2}{2}}
$$

After we plug these back into $\rho_n$, we then simplify the square root terms and also the exponential terms. After these simplifications, we get:
$$
\rho_n \approx \sqrt{\frac{n-2}{2\pi}}
$$

Finally we divide this expression by $\sqrt{n}$ and by taking the limit for $n \to \infty$, we obtain:
$$
\frac{\rho_n}{\sqrt{n}} \approx 
\sqrt{\frac{n-2}{2 \pi n}}
= \sqrt{\frac{1}{2\pi} \cdot \frac{n-2}{n}}
\;\;\stackrel{n \to \infty}{\longrightarrow}\;\;
\sqrt{\frac{1}{2\pi}} 
= \frac{1}{\sqrt{2\pi}}.
$$

The plot confirms this, as we can observe convergence of $\frac{\rho_n}{\sqrt{n}}$ towards the red line, which represents $\frac{1}{\sqrt{2\pi}}$.

# 3. Weak law of large numbers (1.5 points)

The law of large numbers states that if you repeat an experiment independently a large number of times and average the result, what you obtain should be close to the expected value. Formally, the weak law of large numbers states that if 
$X_1, X_2, \ldots X_n$ are independently and identically 
distributed (iid) random variables with a finite expected value
$E(X_i)=\mu<\infty$, then, for any $\epsilon > 0$
$\lim_{n\rightarrow \infty} P(|\bar X_n-\mu|>\epsilon)=0$
where $\bar X_n=\frac{X_1+X_2+\ldots+X_n}{n}$ is the sample mean (the sample mean converges in probability to the expected value).


## a. 

- In R, simulate rolling a fair 6-sided die $n=10\,000$ times using vectorized functions. 
- After each roll, compute the cumulative mean of all previous rolls. 

- Plot the running average against the number of rolls and comment on the result.

## b. 
- Replicate the die-rolling experiment 
 $M=50$ times using a for loop. 
This will result in $50$
paths which you can plot against the number of rolls.

- Plot all 50 paths on the same graph to observe the variation in the running averages.


## c.
What is the theoretical expected value of a fair 6-sided die?


## d. 
Comment on how the running mean of the paths behaves relative to the expected value  as $n$ increases.

## e. 

- Modify the code in b. to use a while loop.

- Stop the loop when the proportion of replications where the difference between the running average and the expected value lies outside the interval 
$[-\epsilon,\epsilon]$ is less than 0.05. Use $\epsilon = 0.01$. 

- How high is $n$ at  the end of the loop?

## f. 

- Repeat the analysis in b, but instead of simulating from a 
discrete uniform distribution, simulate from a continuous standard cauchy distribution using `rcauchy()`. The (standard) cauchy distribution arises from  the ratio of two independent 
mean zero (standard) normally distributed random variables.

- Plot the running mean against increasing 
$n$. Comment on whether the weak Law of Large Numbers seems to hold for this distribution.
Explain why or why not.


# 4. Central limit theorem (1.5 points)

The central limit theorem states that, under certain conditions, the sum of a large number of random variables is approximately normal. Consider the case where
$X_1, X_2,\ldots,X_n$ are i.i.d. random variables
with finite expected values ($E(X_i)=\mu<\infty$)
 and variances $(Var(X_i)=\sigma^2<\infty)$.
The central limit theorem states that the cumulative distribution function  (CDF)
of the normalized random variable
$$
Z_n = \frac{\bar X_n - \mu}{\sigma/\sqrt{n}}
$$
converges to the standard normal CDF as $n\rightarrow\infty$.

## a. 
Derive the expected value of the sample mean $\bar X_n=\frac{X_1+X_2+\ldots+X_n}{n}$.
 
## b. 
Derive the variance of the sample mean $\bar X_n=\frac{X_1+X_2+\ldots+X_n}{n}$.

## c. 
The central limit theorem makes few assumption on the distribution of the random variables $X_i$.
Consider again the die-rolling experiment where $X_i$
is the random variable giving the number in roll $i$.

- Simulate Die Rolls: For each value of $n$ (where 
$n=\{100,1000,10000\}$), simulate rolling a die 
$n$ times.
Record the mean of the $n$ rolls.

- Repeat the above process $M=100$ times for each value of
$n$. This will give you 100 sample means for each $n$.

- Plot a histogram of the 100 sample means for each value of $n$. Observe the shape of the histograms.

- Discuss how the distribution of the sample means changes as $n$ increases. Comment on whether the histograms appear to approach a normal distribution as $n$, illustrating the Central Limit Theorem.


## d. 
Compute expected value $\mu$ and the variance  $\sigma^2$ for the random variable $X_i$ (giving the number in die-roll $i$), which follows a discrete uniform distribution.

## e.
Now that you computed $\mu$ and $\sigma^2$, 
consider $Z_n$.

- For each value of  $n$ (where $n=\{100,1000,10000\}$)
 and for 
$M=\{100,1000,10000\}$ replications, simulate the rolling of $n$ dice and calculate the $Z_n$.

- Plot histograms of $Z_n$ for each combination of 
$n$ and  $M$.

- Discuss how the distribution of $Z_n$
changes as  $n$  and  $M$ increase.
Comment on whether the distributions of 
$Z_n$ appear to approach a normal distribution, illustrating the Central Limit Theorem.

# 5. Readable and efficient code (1 points)

Read over the code below. 

## a. 
Explain (in text) what the code does.

## b.
Explain (in text) what you would change to make the code more readable.

## c.
Change the code according to b. and wrap it in a function. This function should have at most 10 lines (without adding commands to more lines such as `x <- 1; y <- 2`. Such commands will count as 2 lines!). Check that the function called on the same input outputs the same as the provided code.

```{r eval=FALSE}
set.seed(1)
x <- rnorm(100)
z <- rnorm(100)
if (sum(x >= .001) < 1) {
  stop("step 1 requires 1 observation(s) with value >= .001")
}
fit <- lm(x ~ z)
r <- fit$residuals
x <- sin(r) + .01
if (sum(x >= .002) < 2) {
  stop("step 2 requires 2 observation(s) with value >= .002")
}
fit <- lm(x ~ z)
r <- fit$residuals
x <- 2 * sin(r) + .02
if (sum(x >= .003) < 3) {
  stop("step 3 requires 3 observation(s) with value >= .003")
}
fit <- lm(x ~ z)
r <- fit$residuals
x <- 3 * sin(r) + .03
if (sum(x >= .004) < 4) {
  stop("step 4 requires 4 observation(s) with value >= .004")
}
fit <- lm(x ~ z)
r <- fit$residuals
x <- 4 * sin(r) + .04
x
```

```{r eval=FALSE}
set.seed(1) 
x <- rnorm(1000)
y <- 2 + x + rnorm(1000)
df <- data.frame(x, y)

cat("Step", 1, "\n")
fit1 <- lm(y ~ x, data = df[-(1:250),])
p1 <- predict(fit1, newdata = df[(1:250),])
r <- sqrt(mean((p1 - df[(1:250),"y"])^2))  

cat("Step", 2, "\n")
fit2 <- lm(y ~ x, data = df[-(251:500),])
p2 <- predict(fit2, newdata = df[(251:500),])
r <- c(r, sqrt(mean((p2 - df[(251:500),"y"])^2)))

cat("Step", 3, "\n")
fit3 <- lm(y ~ x, data = df[-(501:750),])
p3 <- predict(fit3, newdata = df[(501:750),])
r <- c(r, sqrt(mean((p3 - df[(501:750),"y"])^2)))

cat("Step", 4, "\n")
fit4 <- lm(y ~ x, data = df[-(751:1000),])
p4 <- predict(fit4, newdata = df[(751:1000),])
r <- c(r, sqrt(mean((p4 - df[(751:1000),"y"])^2)))  
r
```


# 6. Measuring and improving performance (2 points)

Have a look at the code of the function below. It is a function for performing a Kruskal-Wallis test, a robust non-parametric method for testing whether samples come from the same distribution. (Note: we assume no missing values are present in `x`). 
```{r}
kwtest <- function (x, g, ...) 
{
    if (is.list(x)) {
        if (length(x) < 2L) 
            stop("'x' must be a list with at least 2 elements")
        if (!missing(g)) 
            warning("'x' is a list, so ignoring argument 'g'")
        if (!all(sapply(x, is.numeric))) 
            warning("some elements of 'x' are not numeric and will be coerced to numeric")
        k <- length(x)
        l <- lengths(x)
        if (any(l == 0L)) 
            stop("all groups must contain data")
        g <- factor(rep.int(seq_len(k), l))
        x <- unlist(x)
    }
    else {
        if (length(x) != length(g)) 
            stop("'x' and 'g' must have the same length")
        g <- factor(g)
        k <- nlevels(g)
        if (k < 2L) 
            stop("all observations are in the same group")
    }
    n <- length(x)
    if (n < 2L) 
        stop("not enough observations")
    r <- rank(x)
    TIES <- table(x)
    STATISTIC <- sum(tapply(r, g, sum)^2/tapply(r, g, length))
    STATISTIC <- ((12 * STATISTIC/(n * (n + 1)) - 3 * (n + 1))/(1 - 
        sum(TIES^3 - TIES)/(n^3 - n)))
    PARAMETER <- k - 1L
    PVAL <- pchisq(STATISTIC, PARAMETER, lower.tail = FALSE)
    names(STATISTIC) <- "Kruskal-Wallis chi-squared"
    names(PARAMETER) <- "df"
    RVAL <- list(statistic = STATISTIC, parameter = PARAMETER, 
        p.value = PVAL, method = "Kruskal-Wallis rank sum test")
    return(RVAL)
}
```

## a. 
Write a pseudo code outlining what the function does.

## b. 
For example data, call the function in two ways: once using `x` as a list and once using `x` as a vector with a corresponding `g` argument. Ensure that the two different function calls return the same thing by aligning the inputs.

## c. 
Make a faster version of `kwtest()` that only computes the Kruskal-Wallis  **test statistic** when the input is a numeric variable `x` and a variable `g` which gives the group membership. You can try simplifying the function above or by coding from the mathematical definition (see https://en.wikipedia.org/wiki/Kruskal%E2%80%93Wallis_one-way_analysis_of_variance). This function should also perform some checks to ensure the correctness of the inputs (use `kwtest()` as inspiration).

## d. 
Consider the following scenario. You have samples available from multiple experiments $m=1000$ where you collect the numerical values for the quantity of interest `x` and the group membership for $n$ individuals. The first $20$ individuals in each sample belong to group 1, the following $20$ individuals in each sample belong to group 2, the last $10$ individuals in each sample belong to group 3. Use the following code to simulate such a data structure:
```{r}
set.seed(1234)
m <- 1000 # number of repetitions
n <- 50   # number of individuals
X <- matrix(rt(m * n, df = 10), nrow = m)
grp <- rep(1:3, c(20, 20, 10))
```
  Write a function which performs the Kruskal-Wallis test using the function             `stats:::kruskal.test.default()` for $m$ repeated experiments and returns a vector of $m$ test statistics. The input of this function are a matrix `X` with $m$ rows and $n$ columns and a vector `g` which gives the grouping. 


## e. 
Write a function which performs the Kruskal-Wallis test using the function in point c. for $m$ repeated experiments and returns a vector of $m$ test statistics.
The input of this function are a matrix `X` with $m$ rows and $n$ columns and a vector `g` which gives the grouping. 

## f. 
Compare the performance of the two approaches using a benchmarking package on the data generated above. Comment on the results.

## g. 
Now consider vectorizing the function in point c. Compare this approach to the other two.
Comment on the results.

